---
output:
  github_document:
    html_preview: false
---


## The Cosmere - Book Length Viz

The [Cosmere](https://coppermind.net/wiki/Cosmere) is a epic fantasy
fictional universe designed by the author
[Brandon Sanderson](https://www.brandonsanderson.com/).

The author has revelead that he plans to write more than 
[40 novels](https://www.reddit.com/r/Stormlight_Archive/comments/5j8bkb/no_spoilersstate_of_the_sanderson_2016/dbgizmb/).
Currently only [23](https://coppermind.net/wiki/Cosmere) books are published.

## Visualisation

In this visualization I want to represent reading time for each of the books
plus total number of words (which is the same thing) but grouped for each
independent Series.

There's also some standalone categories for the books that are not part of any
saga yet.

## Data

I gathered the data mainly from [here]() but also other sources.

All the data is available in this public [gsheet](https://docs.google.com/spreadsheets/d/1vi3ZIA-aka0meB8rOLKByZWHnci1ow2kQ0eez35E_T4/edit?usp=sharing)
document.

## Usage

To use, just run inside R:

```{r cosmere, fig.height=6, fig.path='figures/figure', fig.width=14, message=FALSE, warning=FALSE, dpi=300}
source("script.R")
```


<details>
<summary>You will need to install the following <bold>dependencies</bold> first:</summary>
<br>

```{r results = "asis", echo=FALSE}
glue::glue_data(renv::dependencies("script.R",quiet = T),"* {Package}\n")
```

</details>


